Reanalysis Step 7: Missingness Pattern Analysis

### 2025-09-01 reanalysis-7-missingness_patterns.qmd
fix ai idiocy

# Extract SMDs for plotting
smd_raw <- ExtractSmd(table1)

# ExtractSmd returns a list - extract just the numeric SMD values
# The list has one element per variable, plus a "Variable" element with names
if (is.list(smd_raw) && "Variable" %in% names(smd_raw)) {
  # Get variable names from the Variable element
  var_names <- smd_raw$Variable
  # Get SMD values (all elements except Variable)
  smd_values <- unlist(smd_raw[names(smd_raw) != "Variable"])

  smd_data <- data.frame(
    Variable = var_names,
    SMD = smd_values,
    stringsAsFactors = FALSE
  )
} else if (is.data.frame(smd_raw)) {
  # If it's a data frame, use rownames as variable names
  smd_data <- data.frame(
    Variable = rownames(smd_raw),
    SMD = smd_raw[, 1],
    stringsAsFactors = FALSE
  )
} else {
  # Fallback: use available_vars as variable names
  smd_values <- unlist(smd_raw)
  smd_values <- smd_values[!is.na(suppressWarnings(as.numeric(smd_values)))]

  smd_data <- data.frame(
    Variable = available_vars[1:length(smd_values)],
    SMD = as.numeric(smd_values),
    stringsAsFactors = FALSE
  )
}

# Remove NA values
smd_data <- smd_data[!is.na(smd_data$SMD), ]

# Order by absolute SMD
smd_data <- smd_data[order(abs(smd_data$SMD), decreasing = TRUE), ]

@@

# Plot 1: SMD plot
p1 <- ggplot(smd_data, aes(x = reorder(Variable, abs(SMD)), y = SMD)) +
  geom_point(size = 3, color = ifelse(abs(smd_data$SMD) > 0.1, "red", "blue")) +
  geom_hline(yintercept = c(-0.1, 0, 0.1), linetype = c(2, 1, 2), alpha = 0.5) +
  coord_flip() +
  labs(title = "Standardized Mean Differences: Excluded vs Included",
       subtitle = "Red points indicate SMD > 0.1 (substantial difference)",
       x = "Variable",
       y = "Standardized Mean Difference") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

>>>>

# Extract SMDs with more detail for factor variables
# Get the table itself to access all SMD information
table1_matrix <- print(table1, smd = TRUE, printToggle = FALSE)

# Extract SMDs manually from the table for each row
# This will give us SMDs for each level of factor variables
smd_data <- data.frame()

# Process each variable in the table
for (var in available_vars) {
  if (var %in% names(original_data)) {
    if (var %in% factor_vars && is.factor(original_data[[var]])) {
      # For factor variables, calculate SMD for each level
      var_levels <- levels(original_data[[var]])
      for (level in var_levels) {
        # Calculate proportion in each group
        prop_excluded <- mean(original_data[[var]][original_data$excluded] == level, na.rm = TRUE)
        prop_included <- mean(original_data[[var]][!original_data$excluded] == level, na.rm = TRUE)

        # Calculate SMD for this level
        # SMD for proportions = (p1 - p2) / sqrt((p1*(1-p1) + p2*(1-p2))/2)
        pooled_var <- (prop_excluded * (1 - prop_excluded) + prop_included * (1 - prop_included)) / 2
        smd_val <- if (pooled_var > 0) {
          abs(prop_excluded - prop_included) / sqrt(pooled_var)
        } else {
          0
        }

        smd_data <- rbind(smd_data, data.frame(
          Variable = paste0(var, ": ", level),
          SMD = smd_val,
          VarType = "Factor",
          BaseVar = var,
          stringsAsFactors = FALSE
        ))
      }
    } else {
      # For continuous variables, calculate single SMD
      if (sum(!is.na(original_data[[var]])) > 0) {
        mean_excluded <- mean(original_data[[var]][original_data$excluded], na.rm = TRUE)
        mean_included <- mean(original_data[[var]][!original_data$excluded], na.rm = TRUE)
        sd_pooled <- sqrt((var(original_data[[var]][original_data$excluded], na.rm = TRUE) + 
                          var(original_data[[var]][!original_data$excluded], na.rm = TRUE)) / 2)

        smd_val <- if (!is.na(sd_pooled) && sd_pooled > 0) {
          abs(mean_excluded - mean_included) / sd_pooled
        } else {
          0
        }

        smd_data <- rbind(smd_data, data.frame(
          Variable = var,
          SMD = smd_val,
          VarType = "Continuous",
          BaseVar = var,
          stringsAsFactors = FALSE
        ))
      }
    }
  }
}

# Remove NA values and order by absolute SMD
smd_data <- smd_data[!is.na(smd_data$SMD), ]
smd_data <- smd_data[order(abs(smd_data$SMD), decreasing = TRUE), ]

# Get top differences (for display purposes, limit to top 20)
smd_data_top <- head(smd_data, 20)

@@

# Plot 1: SMD plot (showing top 20 differences)
p1 <- ggplot(smd_data_top, aes(x = reorder(Variable, abs(SMD)), y = SMD)) +
  geom_point(aes(color = VarType), size = 3) +
  geom_point(data = smd_data_top[abs(smd_data_top$SMD) > 0.1, ],
             color = "red", size = 3.5) +
  geom_hline(yintercept = c(-0.1, 0, 0.1), linetype = c(2, 1, 2), alpha = 0.5) +
  coord_flip() +
  labs(title = "Standardized Mean Differences: Excluded vs Included (Top 20)",
       subtitle = "Red overlay indicates SMD > 0.1; Factor levels shown separately",
       x = "Variable",
       y = "Standardized Mean Difference",
       color = "Variable Type") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
